---
title: "Stats506_Project"
author: "Heleyna Tucker"
format:
  html:
    embed-resources: true
editor: visual
---

# Project - Light data in buildings

### Load in data:

```{r}
library(tidyverse)
energy <- read_csv('/Users/19892/Downloads/cbecs2018_final_public.csv')
engery_comp <- energy[c("LTZHRP", "LTOHRP", "LTNHRP", "FLUORP", "CFLRP", "BULBP", "HALOP", "HIDP", "LEDP", "LEDP", "OTLTP"),]
```

Read in File below and get necessary data:

```{r}
energy <- read_csv('/Users/19892/Downloads/cbecs2018_final_public.csv')
energy_comp <- energy[c("PUBID","LTZHRP", "LTOHRP", "LTNHRP", "FLUORP", "CFLRP", "BULBP", "HALOP", "HIDP", "LEDP", "OTLTP")]
head(energy_comp)
```

```{r}
#Rename columns to be easily identifiable
colnames(energy_comp) <- c("buildingID", "pLitNoHours", "pLitOpen", "pLitOffHours", "pFluor", "pCompFluor", "pIncandes", "pHalogen", "pHID", "pLED", "pOther")
head(energy_comp)
```

For the sake of data analysis, I will assume that the NA values are 0 percentage values:

```{r}
energy_comp[is.na(energy_comp)] <- 0

#Add varivale to total the percentage of lights in each building:
energy_comp["pTotalLights"] <- energy_comp$pFluor + energy_comp$pCompFluor + energy_comp$pIncandes + energy_comp$pHalogen + energy_comp$pHID + energy_comp$pLED + energy_comp$pOther

```

Because some totals add up to over/under 100% or 0% which means there was some inconsistency in the data collection for this dataset. For the sake of this project, I will remove these values from the data we will be doing the analysis on:

```{r}
length(which(energy_comp$pTotalLights ==100))
length(energy_comp$pTotalLights)
```

As we can see above, there is about 80% of the data that has the percentage of lights that add up to 100%, still giving us a good amount of data to work with.

```{r}
#Remove rows of data that have pTotalLights not equal to 100
energy_comp <- energy_comp[energy_comp$pTotalLights == 100, ]
dim(energy_comp)
head(energy_comp)
```

Look at a Summary:

```{r}
summary(energy_comp)
```

### Preliminary Graphs:

```{r}
correlation_matrix <- cor(energy_comp[, c("pLitOpen", "pLitOffHours", "pLitNoHours", "pFluor", "pCompFluor", "pIncandes", "pHalogen", "pHID", "pLED", "pOther")])
correlation_matrix

heatmap(correlation_matrix, 
        col = colorRampPalette(c("blue", "white", "red"))(100),
        scale = "column",  # Add a legend for columns
        main = "Correlation Heatmap with Legend"
)
```

Not many correlation differences in variables interested in.

### Try KMeans clustering:

Elbow Plot is graphed below:

```{r}
wcss <- vector("numeric", length = 15)  # Set the maximum number of clusters you want to consider

for (i in 1:15) {
  kmeans_result <- kmeans(energy_comp[, c("pFluor", "pCompFluor", "pIncandes", "pHalogen", "pHID", "pLED", "pOther")], centers = i)
  wcss[i] <- kmeans_result$tot.withinss
}
```

```{r}
k_values <- 1:15  # Assuming you considered up to 10 clusters

ggplot() +
  geom_line(aes(x = k_values, y = wcss), color = "blue") +
  geom_point(aes(x = k_values, y = wcss), color = "red") +
  labs(title = "Elbow Method", x = "Number of Clusters (k)", y = "Within-Cluster Sum of Squares (WCSS)")

```

About k=4 is where the elbow is of the plot. Using k = 4 in kmeans clustering below:

```{r}
# Assuming df is your data frame with lighting percentages
set.seed(123)  # for reproducibility
k <- 4  # specify the number of clusters
kmeans_result <- kmeans(energy_comp[, c("pLitOpen", "pLitOffHours", "pLitNoHours")], centers = k)
energy_comp$cluster <- as.factor(kmeans_result$cluster)
#Visualize first couple rows of the clustered data:
head(energy_comp)
```

How many values in each cluster:

```{r}
n1 <- length(which(energy_comp$cluster == 1))
n2 <- length(which(energy_comp$cluster == 2))
n3 <- length(which(energy_comp$cluster == 3))
n4 <- length(which(energy_comp$cluster == 4))
n1
n2
n3
n4
```

Summary Stats for each Cluster:

```{r}
# Calculate summary statistics for each cluster
cluster_summary <- aggregate(. ~ cluster, data = energy_comp[, c("cluster", "pLitOpen", "pLitOffHours", "pLitNoHours","pFluor", "pCompFluor", "pIncandes", "pHalogen", "pHID", "pLED", "pOther")], mean)

# Print the summary
print(cluster_summary)
```

```{r}
library(tidyr)


# Reshape data for ggplot
data_long <- gather(cluster_summary, key = "light_type", value = "percentage", -cluster)

# Separate data for two bar charts
bar_chart1_data <- data_long[data_long$light_type %in% c("pLitOpen", "pLitOffHours", "pLitNoHours"), ]
bar_chart2_data <- data_long[!data_long$light_type %in% c("pLitOpen", "pLitOffHours", "pLitNoHours"), ]

# Create bar chart for pLitOpen, pLitOffHours, and pLitNoHours with upper error bars
p1 <- ggplot(bar_chart1_data, aes(x = factor(cluster), y = percentage, fill = light_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(aes(ymin = percentage, ymax = percentage + 5), position = position_dodge(width = 0.9), width = 0.25) +  # Show only upper error bars
  labs(title = "Percentage of Lighting Types (pLitOpen, pLitOffHours, pLitNoHours) by Cluster",
       x = "Cluster",
       y = "Percentage") +
  theme_minimal()

# Create bar chart for the rest of the variables with upper error bars
p2 <- ggplot(bar_chart2_data, aes(x = factor(cluster), y = percentage, fill = light_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(aes(ymin = percentage, ymax = percentage + 5), position = position_dodge(width = 0.9), width = 0.25) +  # Show only upper error bars
  labs(title = "Percentage of Other Lighting Types by Cluster",
       x = "Cluster",
       y = "Percentage") +
  theme_minimal()

# Display the two plots separately
p1
p2

```

two plots and error bars.
